%
%  chapter2.tex
%
\chapter{State of the Art}
\label{chap:stateofart}

This chapter surveys the technical and research landscape relevant to U!REKA Play4Change. It covers gamification in education, serious games, multiplatform mobile development, AI-generated content, semantic deduplication, and related platforms. The chapter concludes by identifying the gap that the platform addresses.

\section{Gamification in Education}
\label{sec:gamification}

Gamification is defined as the application of game design elements in non-game contexts \cite{deterding2011gamification}. The term distinguishes the application of specific game mechanics --- points, leaderboards, badges, progress bars, and challenge structures --- from full-scale game design. In educational settings, gamification is applied to existing content and activities with the goal of increasing learner motivation without changing the learning material itself.

The underlying motivation rests on a well-established observation in motivational psychology: game mechanics activate intrinsic drivers of engagement that are otherwise absent in purely didactic settings. Competition, mastery progression, and immediate reward all contribute to sustained voluntary effort. Gamification attempts to transfer these drivers into learning contexts where they do not arise naturally.

\subsection{Evidence of Effectiveness}

Empirical evidence on gamification's effectiveness in education has accumulated steadily. A systematic review of 24 empirical studies found positive effects in the majority of cases, with social comparison mechanisms such as leaderboards and peer challenges producing the most consistent engagement gains \cite{hamari2014does}. The review analysed studies across a range of learning contexts, student ages, and gamification element types, finding that point systems and progress indicators are the most consistently effective elements across diverse populations.

However, the same review noted important moderating factors: the specific gamification element employed, the duration of the intervention, and the demographic profile of learners all influence outcomes. Long-duration interventions show diminishing effects if the novelty of the game element wears off and is not replaced by genuine mastery progression. This finding informs the design of U!REKA Play4Change's scoring engine, which combines individual progression with configurable penalty and reward thresholds rather than relying solely on competitive ranking, and sustains novelty through AI-generated content that changes daily.

\section{Serious Games}
\label{sec:seriousgames}

Serious games are digital games designed with a primary purpose beyond pure entertainment, typically targeting learning, training, or behaviour change \cite{zhonggen2019meta}. Unlike gamification, which adds game elements to an existing non-game activity, a serious game is a game in its own right --- it has goals, rules, feedback mechanisms, and voluntary participation --- but its design is oriented around a learning or behavioural objective.

A meta-analysis of serious game research spanning a decade found statistically significant improvements in learning outcomes across multiple disciplines, including science, health, and civic education \cite{zhonggen2019meta}. The analysis identified two mechanisms as the strongest predictors of learning gain: immediate feedback after every action, and clear goal structures that give players an unambiguous sense of what they are working toward. Feedback enables learners to adjust understanding in real time; goal structure provides the motivational frame that makes continued effort rational.

U!REKA Play4Change draws directly on these findings. Each daily task is structured as a self-contained challenge with a defined correct response, delivering immediate scoring feedback. The daily goal --- completing the day's task --- is unambiguous. The streak and penalty system extends the goal structure across days, making each day's engagement part of a larger progression narrative.

\section{Multiplatform Mobile Development}
\label{sec:multiplatform}

\subsection{Native versus Cross-Platform Approaches}

Mobile application development historically bifurcated into two models. The native model uses separate codebases per platform: Swift or Objective-C for iOS, and Java or Kotlin for Android. Each codebase is developed, tested, and maintained independently, with full access to each platform's APIs and the highest achievable performance. The cross-platform model uses a single codebase that targets multiple platforms through an abstraction layer, trading some degree of platform-specific control for reduced duplication.

The native model's engineering cost is its decisive weakness for small teams: any feature, bug fix, or design change must be implemented twice, in two different languages, and verified on two different platforms. At scale, this doubles the engineering resource required for any given throughput of feature delivery.

Established cross-platform frameworks each make different architectural choices. React Native compiles JavaScript logic to native platform components but routes all interaction through a JavaScript bridge, introducing latency that manifests in complex animated interfaces. Flutter compiles to native ARM code and provides its own rendering engine, achieving near-native performance but drawing all UI components itself rather than using platform-native widgets, which can diverge from each platform's visual conventions. KMP (Kotlin Multiplatform) takes a fundamentally different approach: it shares only business logic across platforms while allowing the user interface layer to use either platform-native frameworks or the shared Compose Multiplatform declarative UI toolkit. This means KMP inherits native platform UI components where desired and shares UI code where appropriate, achieving a balance that neither React Native nor Flutter offers.

\subsection{Justification for Kotlin Multiplatform}

U!REKA Play4Change adopts KMP for shared business logic --- including the scoring engine, data models, network clients, and local caching strategies --- combined with Compose Multiplatform for the shared declarative user interface. This combination allows Android and iOS to share both logic and UI components from a single codebase while maintaining access to native platform APIs where required. The choice reduces duplication and accelerates feature delivery without sacrificing the performance characteristics required for smooth animated game interfaces.

The decision was supported by the team composition: the primary developer's existing expertise is in Kotlin for Android. KMP allows the same language and the same idioms to be extended to iOS rather than requiring the acquisition of a second language and platform toolchain. This pragmatic alignment of technology choice with team capability is itself an engineering consideration: a solution that the team can implement correctly is preferable to a theoretically superior solution that the team cannot.

\section{AI-Generated Content in Education}
\label{sec:aicontent}

\subsection{Large Language Models for Educational Content}

Large language models (LLMs) are AI systems trained on large corpora of text to predict and generate coherent language. Their application to educational content generation is a rapidly growing area of research and practice. LLMs can produce question-and-answer pairs, explanatory summaries, and scenario-based tasks at scale and at low marginal cost compared to human authoring, making them an attractive solution for the content authoring bottleneck identified in Chapter~\ref{chap:introduction}.

The primary engineering challenges when using LLMs for educational content generation are: factual accuracy, since models can produce plausible but incorrect statements; content diversity, since models tend to cluster around common framings of a topic; structural consistency, since models may produce responses that do not conform to the required schema; and deduplication, since models have no memory between invocations and will regenerate conceptually equivalent content given sufficiently similar prompts. Each of these challenges requires a deliberate pipeline design rather than naive invocation of the model API.

\subsection{LangChain4j and the Mistral API}

LangChain4j is a Java library that provides abstractions for building LLM-powered applications, including prompt templating, model chaining, output parsing, and schema validation \cite{langchain4j2024}. It supports multiple model backends through a uniform interface, enabling the underlying model to be replaced without changing application code. U!REKA Play4Change uses LangChain4j to orchestrate nightly batch jobs that generate daily tasks parameterised by subject domain and difficulty level, validate the structure of each response against a defined schema, and retry failed validations before discarding malformed outputs.

The Mistral API provides access to a family of instruction-tuned language models well suited to structured text generation tasks \cite{mistral2024}. Instruction-tuned models are trained to follow explicit instructions embedded in the prompt rather than simply completing the text, making them more reliable for constrained generation tasks such as producing valid JSON-structured educational tasks. The combination of LangChain4j's structured output capabilities with Mistral's instruction-following behaviour provides the reliability required for an automated nightly pipeline.

\subsection{Semantic Deduplication with pgvector}

A fundamental property of language models is that they have no memory between separate invocations. A model asked today to generate a sustainability task and asked again tomorrow has no recollection of yesterday's output. As the task library grows, the probability that a newly generated task is conceptually equivalent to an existing one increases. Crucially, equivalent tasks are paraphrases, not duplicates: "sort your recycling correctly" and "separate your waste into the appropriate bins" convey the same educational action with no overlapping words. Exact text comparison cannot detect this equivalence.

The correct approach is to compare the meaning of texts rather than their words. Vector embeddings are numerical representations of text meaning: every piece of text is mapped to a point in a high-dimensional space, where texts with similar meanings are mapped to nearby points. A sentence embedding model --- a model trained specifically to produce embeddings where semantic similarity corresponds to geometric proximity --- converts each generated task into a vector before storage.

The geometric metric used for comparison matters. Euclidean distance measures the straight-line distance between two points in the embedding space. This distance is sensitive to the magnitude of the vectors, which is influenced by sentence length rather than meaning. Cosine similarity measures the angle between two vectors rather than the distance between their tips. Because the angle is independent of vector magnitude, it correctly captures semantic similarity regardless of sentence length differences. Cosine similarity is the standard metric for semantic text comparison and is the metric used by pgvector \cite{pgvector2024}.

PostgreSQL extended with the pgvector module enables vector storage and cosine similarity queries natively, without a separate vector database system. This integration means that semantic deduplication queries and relational data queries share a single transactional context, eliminating consistency risks that arise when data is distributed across multiple storage systems. An approximate nearest-neighbour index reduces the query cost from linear in the number of stored tasks to approximately sub-linear, making the deduplication check tractable as the task library grows.

\section{Related Platforms}
\label{sec:relatedplatforms}

Several gamified learning platforms have gained widespread adoption and serve as the primary reference points for this work.

Duolingo applies a streak-based engagement model to language learning and is the most studied platform for the long-term effectiveness of gamification \cite{hamari2014does}. Its retention mechanisms --- streaks, leaderboards, and daily goals --- are among the best validated in the literature. However, all of Duolingo's content is hand-authored by domain specialists. This model produces high-quality content but cannot be replicated by organisations without Duolingo's scale and budget. Duolingo is also a single-domain platform; its content model does not generalise. The platform delivers native Android and iOS applications built from separate native codebases.

Kahoot is a synchronous, instructor-led quiz platform designed for classroom use \cite{hamari2014does}. Players join a live session hosted by an instructor and answer questions on their own devices while results are displayed publicly. This model is effective for classroom engagement but is not suitable for self-paced asynchronous learning, which is the target use case for U!REKA Play4Change. Kahoot does not use AI-generated content.

Quizlet is a flashcard and study tool platform that allows communities to create and share content. It provides no scoring engine, no AI-generated content, and no streak or penalty mechanics. Its community content model distributes authoring effort across users but provides no quality control and no deduplication. Quizlet's mobile clients are built as separate native applications.

None of these platforms combines automated AI content generation with multiplatform delivery from a single shared codebase.

\section{Identified Gap}
\label{sec:gap}

The survey above reveals a consistent pattern: the platforms that have achieved proven engagement outcomes rely on manually authored content, and the platforms that support automated content generation do not apply the engagement mechanics and delivery infrastructure that sustain long-term learner motivation. Furthermore, no existing platform in this space is known to deliver both Android and iOS applications from a single shared codebase while also solving the AI content generation and deduplication problem.

U!REKA Play4Change addresses this gap by combining four capabilities that have not previously been integrated in a single platform: multiplatform delivery from a single KMP codebase; fully automated nightly AI content generation via LangChain4j and the Mistral API; semantic deduplication via pgvector to prevent content fatigue; and a server-side scoring engine with reward and penalty mechanics. Each of these capabilities exists independently in the literature and in commercial products. Their combination in a single deployable system is the novel contribution of this work.
